--- 
title: "Handbook of Regression Modeling in People Analytics"
subtitle: "With Examples in R and Python"
author: "Keith McNulty"
site: bookdown::bookdown_site
always_allow_html: true
documentclass: krantz
bibliography: r/book.bib
fig-caption: yes
link-citations: yes
github-repo: keithmcnulty/peopleanalytics-regression-book
---

```{r, echo = FALSE}
library(reticulate)
matplotlib <- import("matplotlib")
matplotlib$use("Agg", force = TRUE)

if (knitr::is_latex_output()) {
  options(width = 100)
}
```


`r if (knitr::is_latex_output()) '<!--'`
# Welcome {-}

Welcome to the website for the book *Handbook of Regression Modeling in People Analytics* by [Keith McNulty](https://twitter.com/dr_keithmcnulty).  Content can be used freely with appropriate referencing.

This book is available in [bootstrap format](http://peopleanalytics-regression-book.org) and, for those who prefer, in a more plain [gitbook format](http://peopleanalytics-regression-book.org/gitbook).
`r if (knitr::is_latex_output()) '--> \\newpage\\thispagestyle{empty}\\null'`

### Notes on data used in this book {-}

For R and Python users, each of the data sets used in this book can be downloaded individually by following the code in each chapter.  Alternatively for R users who intend to work through all of the chapters, all data sets can be loaded into an R session in advance by installing and loading the `peopleanalyticsdata` R package.  

```{r, eval = FALSE}
# install and load all data sets in the book
install.packages("peopleanalyticsdata")
library(peopleanalyticsdata)

# see a list of data sets
data(package = "peopleanalyticsdata")

# find out more about a specific data set ('managers' example)
help(managers)
```

`r if (knitr::is_latex_output()) '<!--'`
### Comments on this book {-}

This book is in permanent open review.  Feedback and requests for new content should be made via the [issues](https://github.com/keithmcnulty/peopleanalytics-regression-book/issues) section of the book's Github repo.

### Solutions to exercises {-}

It is not my intention to publish a comprehensive set of solutions to the exercises in this book.  Many of the exercises can be approached in different ways and I think it is important that readers apply their learning without being constrained to an example solution.  However, some specific questions have generated considerable interest from readers and I have started to post example solutions to those questions [here](https://keithmcnulty.github.io/peopleanalytics-regression-book/solutions/).  Illustrative solutions submitted by readers will progressively be posted there also.  


Happy modeling!

*Last update:  `r format(Sys.Date(), "%d %B %Y")`*

`r if (knitr::is_latex_output()) '-->'`







