---
title: "Chapter 3 Data Exercise 9 Solution"
author: "Keith McNulty"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Question:** By using an appropriate hypothesis test, determine if the age of those who have recently donated is at least 10 years older than those who have not recently donated in the population.

**Solution:** Let $x_1$ be the vector of ages of those who have recently donated and let $x_0$ be the vector of ages of those who have not recently donated. 

```{r}
# get data
url <- "http://peopleanalytics-regression-book.org/data/charity_donation.csv"
donation <- read.csv(url)

# subset ages for those who recently donated and those who didn't
x1 <- subset(donation, subset = recent_donation == 1, select = "age")
x0 <- subset(donation, subset = recent_donation == 0, select = "age")
```

We are trying to establish if $\bar{x_1} - 10 > \bar{x_0}$ in the population. Alternatively stated, we are testing if $\overline{x_1 - 10} > \bar{x_0}$ in the population, so we are doing a $t$-test in our sample to compare `x1 - 10` with `x0`.  Our null hypothesis is that there is no difference between the means of `x1 - 10` and `x0`.  Our alternative hypothesis is that the mean of `x1 - 10` is greater than the mean of `x0`.  This requires a one sided $t$-test.

```{r}
# run one sided t-test to compare means of x1 - 10 and x0
t.test(x1 - 10, x0, alternative = "greater")
```

The p-value of this test meets an alpha standard of 0.001 and this confirms the alternate hypothesis that those who donated are at least ten years older than those who did not.

[Back to solutions](index.html)
