---
title: "Modeling Ordered Ratings using Proportional Odds Regression"
author: "Keith McNulty"
output:
  xaringan::moon_reader:
    css: style.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

class: left, middle, r-logo

## Note on source

This document is a summary learning developed for the NY Strategic HR Analytics Meetup on 15 January 2021.  It is based on material in the open source book *[Handbook for Regression Modeling in People Analytics](http://peopleanalytics-regression-book.org)*. Please consult this book for a deeper understanding/treatment. 

## Note on languages

This document is coded in R.  The model type used in this document (proportional odds logistic regression) is not currently well supported in Python.  It will become available in the next release of the `statsmodels` package.


---
class: left, middle, r-logo

## Example data set:  salespeople performance ratings

Let's look at some data that relates to the performance ratings of sales people in an organization on a scale of 1 to 5 of increasing performance:

```{r}
## download data
url <- "https://raw.githubusercontent.com/keithmcnulty/peopleanalytics-regression-book/master/presentations/employee_performance.csv"

employee_performance <- read.csv(url)

## look at the data
head(employee_performance)
```

We are being asked which of the other four factors influence the performance rating and how do they influence it.

---
class: left, middle, r-logo

## What does our data look like?

```{r}
## look at the data type - are they appropriate?
str(employee_performance)
```

---
class: left, middle, r-logo

## Adjusting data types for appropriate modeling

We need to make sure that `region` and `gender` is understood as an **unordered** categorical column of data.  We also need to make sure that the `rating` column is understood as an **ordered** categorical column.

```{r}
## unordered categories
cat_columns <- c("region", "gender")
employee_performance[cat_columns] <- 
  lapply(employee_performance[cat_columns], 
         as.factor)

## ordered categories
employee_performance$rating <- 
  ordered(employee_performance$rating,
          levels = 1:5)

str(employee_performance)
```

---
class: left, middle, r-logo

## Quick visualizations of bivariate relationships in the data

```{r, fig.height=6}
library(GGally)

## create a pairplot
GGally::ggpairs(employee_performance)
```
